\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2017-10-26 document/XSGRB_sample.tex

GRBs are interesting as they provide the opportunity to derive new insights
into a very wide range of topics in astrophysics. In that sense GRB
research is very multidisciplinary. The {\it Swift} satellite \citep{Gehrels09}, 
which was launched in 2004, has made
it possible to harvest the great potential in using GRBs as probes of
galaxies in the intergalactic medium, which was already hinted at by
results from earlier missions.

A crucial aspect of the great success of the {\it Swift} mission has been
the extensive ground based follow-up observations of the afterglows and of
the host galaxies of the GRBs. This work has involved a large community
of researchers both inside and outside the official {\it Swift} team. This fruitful
collaboration has been greatly facilitated by the open data access policy
of the {\it Swift} mission.

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs (typically
R=$\lambda/\Delta\lambda$$<$1000) \citep[e.g.][]{Fynbo2009}. For a handful of
very bright afterglows high-resolution (typically R$>$20000) spectra were
secured.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd generation
instruments at the ESO Very Large Telescope (VLT). It is a high-throughput 
echelle spectrograph
covering the full spectra range from the UV atmospheric cutoff to the near-infrared
(near-IR)
K-band with three spectral arms (UVB, VIS, NIR). It was designed very much with
transient source follow-up in mind as the fading luminosities of such sources 
makes in urgent to secure as extensive coverage as possible in the shortest possible 
time. At the same time the resolution was set to be in the range 4000--9000 in order
to be able to get a large useful spectra range between the many sky-background
emission lines in the red and near-IR.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria and follow-up}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

X-shooter can cover the spectral wavelength region from 3000 \AA~to 24800 \AA~in
a single exposure, by splitting the light in three separate spectroscopic arms
using dichroics. This way each arm work as a separate instrument, each
functioning as its own echelle spectrograph. The ultraviolet blue(UVB) arm
covering 3000 - 5500 \AA, the visual(VIS) arm covering 5500 - 10200 \AA, and the
near-infrared()NIR) arm covering 10200 - 24800 \AA~ with the possibility of
applying a k-band blocking filter cutting the coverage of the NIR arm at 10200 -
21000 \AA. For all observations, a nod-and-shuffle observing scheme has been
employed, with a nodding throw of 5 \arcsec. The individual observations has
been carried out in a standard ABBA pattern, which allows flexibility in the sky
subtraction method for the different arms.

@@

@@

\subsection{GRB120909A (z = 3.929)}
The data presented here has formed the basis of GCN \#
13730\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13730.gcn3}}, but is not
published otherwise. A very rapid follow-up, starting only 1.7 hours after the
BAT trigger, this 1.2 ks observation captures a very bright afterglow
continuum, starting at 4500 \AA, signifying the onset of the Lyman limit for a 
system at $z = 3.929$. Absorption from high-column density hydrogen leaves very
prominent absorption features in the form of \lya, \lyb, and \lyg, visible in
the Lyman alpha forest. Metal absorption lines arising from \feii, \NIii,
\SIii, \sii, \alii, \aliii, \cii, \oi, \civ, and \znii~are all detected along
with the corresponding fine structure lines from (\feii$^*$, \SIii$^*$,
\oi$^*$, \oi$^**$, \cii$^*$), securely anchoring the redshift of the host.

>>>>

Gamma-Ray bursts (GRBs) are interesting as they provide the opportunity to
derive new insights into a very wide range of topics in astrophysics. Examples
range from small scale phenomena relating to the magnetars, properties of highly
relativistic jets, hyper/supernova explosions, the interstellar medium, dust extinction
curves, starburst galaxies, chemical and molecular abundances, escape of ionizing 
radiation, the ionization state of the intergalactic medium, intervening
absoption systems to standard candles in cosmology.  In that sense GRB research
is very multidisciplinary.

The {\it Swift} satellite \citep{Gehrels09}, which was launched in 2004, has
made it possible to harvest much of the great potential in using GRBs as probes
of galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.][]{Ricker04}.  A crucial aspect of the great
success of the {\it Swift} mission has been the extensive ground based
follow-up observations of the afterglows and of the host galaxies of the GRBs.
This work has involved a large community of researchers both inside and outside
the official {\it Swift} team. This fruitful collaboration has been greatly
facilitated by the open data access policy of the {\it Swift} mission.

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs 
\citep[typically
R=$\lambda/\Delta\lambda$$<$1000, e.g.][]{Fynbo2009}. Low-resolution
spectroscopy is powerful for a number of reasons, first of all is the ability
to secure information for even very faint targets. This allows the measurements
of a number of important parameters such as redshifts, spectral slopes,
and extinction. For
a handful of very bright afterglows high-resolution (typically R$>$20000)
spectra were secured and for these events much more information about
conditions inside the host galaxies were extracted 
\citep[e.g.,][]{Fiore,Thoene07,Prochaska07,Vreejswijk07,CastroTirado2010}.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd
generation instruments at the ESO Very Large Telescope (VLT). It is a
high-throughput echelle spectrograph covering the full spectral range from the
UV atmospheric cutoff around 300 nm to the near-infrared (near-IR) K-band 
ending around 2500 nm with three spectral
arms (UVB, VIS, NIR). It was designed very much with transient source follow-up
in mind as the fading luminosities of such sources makes in urgent to secure as
extensive coverage as possible in the shortest possible time. At the same time
the resolution was set to be in the range 4000--9000 in order to be able to get
a large useful spectral range between the many sky-background emission lines in
the red and near-IR spectral ranges. In this paper we present the results
of a dedicated effort over the years 2009 -- 2016 to use the X-shooter
spectrograph to secure spectroscopic observations of
afterglows and host galaxies of GRBs detected by {\it Swift}. 
All data resulting from the survey are publicly availble in reduced form.

The paper is organized in the following way: In Sect.~\ref{sample} we describe
the sample including the sample selection criteria and the observational
strategy.  In Sec.t~\ref{results} we describe the results of the survey, i.e.
the efficiency of the follow-up effort and the characteristics of the observed
bursts. We also here assess the completeness of the finally realized sample.
Finally, we offer our conclusions in Sect.~\ref{conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria and observations}\label{sample}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

X-shooter can cover the spectral wavelength region from 300 nm to 2480 nm in a
single exposure, by splitting the light in three separate spectroscopic arms
using dichroics.  This way each arm work as a separate instrument, each
functioning as its own echelle spectrograph.  The ultraviolet blue(UVB) arm
covering 3000 - 550 nm, the visual(VIS) arm covering 550 - 1020 nm, and the
near-infrared()NIR) arm covering 10200 - 2480 nm with the possibility of
applying a k-band blocking filter cutting the coverage of the NIR arm at 1020 -
2100 nm. For all observations, a nod-and-shuffle observing scheme has been
employed, with a nodding throw of 5 \arcsec. The individual observations has
been carried out in a standard ABBA pattern, which allows flexibility in the sky
subtraction method for the different arms.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion and conclusions}\label{conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The sample we present here include spectroscopic observations of 86 
systems fulfilling our sample criteria of which 72 are afterglow spectra 
and 14 are observations of the underlying host galaxies. 
UPDATE THIS.
All spectra have been made publicly available in the reduced form used
in this paper.

Our sample serves the purpose to further characterize the environments of GRBs
that was also much advanced by the previous surveys based primarily on
lower-resolution spectographs. The most striking result found is that GRBs
typically are located behind very large column densities of neutral hydrogen.
Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ quasars have a handful such systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified \citep[e.g., five
in][]{Noterdaeme2012}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep[][and this work]{Jakobsson2006,
Fynbo2009,Cucchiara2015}.  With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we hence characterize the properties of
star-forming galaxies over cosmic history in terms of redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, extinction curves,
etc.  A number of independent papers have been published or submitted for
publication focusing on many of these specific issues of our sample such as
extinction curves [][see also
\cite{Fynbo2014,Heintz2017}]\citep{Japelj2015,Zafar17}, emission lines from the
underlying host galaxies \citep{Kruhler2015}, the frequency of intervening
\ion{Mg}{ii} absorbers \citep{Christensen2017}, and escape of ionizing
radiation \citep{Tanvir17}.

It seems important to end this paper by stressing that the potential of 
using GRB sightlines as probes is far from fully harvested. The sample of
sightlines probed by our spectra is not representative for all GRB 
sightlines as we have shown and consistent with earlier findings
from samples based on low-resolution spectroscopy \citep[e.g.,][]{Fynbo2009}
and from studies of complete samples of GRB host galaxies 
\citep{Hjorth2012,Covinto2013,Perley2016}. 
\cite{Kruhler2013} convincingly argue, that 
very rich sightlines like that probed by the remarkable GRB\,080607 
\citep{Prochaska09} are probably significantly more frequent than in the sightlines 
sampled by our spectra. However, with current instrumentation these sightlines are
out of reach except under very fortunate circumstances as in the case of 
GRB\,080607 when the afterglow could be observed only a few minutes after the 
burst with a 10-m class telescope. Observations of such sightlines with 
X-shooter-like spectrographs on th enext generation of 20--40-m telescopes
is likely to be very rewarding.

\begin{acknowledgements}
*** We should add an ode to Neil Gehrels here ***.
JPUF, BMJ and DX acknowledge support from the ERC-StG grant EGGS-278202.  The
Dark Cosmology Centre is funded by the Danish National Research Foundation.  TK
acknowledges support by the European Commission under the Marie Curie
Intra-European Fellowship Programme in FP7.  AdUP acknowledges support by the
European Commission under the Marie Curie Career Integration Grant programme
(FP7-PEOPLE-2012-CIG 322307).  This work made use of data supplied by the UK
{\it Swift} Science Data Centre at the University of Leicester.  Finally, we
acknowledge expert support from the ESO staff at the Paranal and La Silla
observatories in obtaining these target of opportunity data.

\end{acknowledgements}

@@

\subsection{GRB120909A (z = 3.929)}
The data presented here has formed the basis of GCN \#
13730\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13730.gcn3}}, but is not
published otherwise. A very rapid follow-up, starting only 1.7 hours after the
BAT trigger, this 1.2 ks observation captures a very bright afterglow
continuum, starting at 450 nm, signifying the onset of the Lyman limit for a 
system at $z = 3.929$. Absorption from high-column density hydrogen leaves very
prominent absorption features in the form of \lya, \lyb, and \lyg, visible in
the Lyman alpha forest. Metal absorption lines arising from \feii, \NIii,
\SIii, \sii, \alii, \aliii, \cii, \oi, \civ, and \znii~are all detected along
with the corresponding fine structure lines from (\feii$^*$, \SIii$^*$,
\oi$^*$, \oi$^**$, \cii$^*$), securely anchoring the redshift of the host.
